Genome Function Phylogenetics

### 2019-04-19 final_project/paper/Paper.Rmd
Final Draft stage

`r newthought("When we build phylogenetic trees,")` we often use the sequences of certain genes as the basis for our tree.
They are the direct substrate of evolution so that makes sense, but when we think of evolution on a macro level, we don't really think of the sequences, we think of phenotype.
Using phenotypes for evolutionary trees is a hard thing to do though because the choice of characteristics can be kind of arbitrary (http://wiki.c2.com/?BorgesClassificationOfAnimals Chinesische Tiere (Jorge Luis Borges 1966)), because it is very far away from the material evolution actually happens on, i.e. DNA.
In this paper, I tried to go a middle ground: Instead of looking at the genetic sequences or the macromolecular phenotype, I'm looking at the presence of certain biological processes, molecular functions, or cellular components, in other words: along the path of evolution, any meaningful change in an organism goes along with the development, change, or loss of such a function; what good is a change in the DNA sequence if it does not lead to a modified function of that gene?

@@

## Data
Annotating genes with their functions can be done experimentally (e.g. by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try and predict the function of a given gene.
Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner (CITE PAPER).
We have been applying this pipeline to whole-genome assemblies of different plant species and generated functional annotations for every gene in each genome.
These annotation sets are (or will be shortly) available from  
\scriptsize https://dill-picl.org/projects/gomap/gomap-datasets \normalsize.

@@

Applying this formula to all pairwise combinations of the genomes we're looking at yields a $S \times S$ distance matrix that can then serve as the input for a neighbor joining algorithm (provided by PHYLIP).

@@

The tree generated by neighbor joining is displayed in Figure \ref{fig:result}.
4 of the 5 maize assemblies are grouped together in a clade, while the last one is an outlier to the remaining grasses and sits on a very long branch.
In the remaining plants, cotton is an outlier to the legumes, which are all grouped in a single clade.

\begin{figure*}[htbp]
\includegraphics{resources/result.png}
\caption{Resulting tree. maize\_v3 and maize\_v4 are two assemblies of the same maize cultivar, medicago1 and medicago2 are two different genotypes.}
\label{fig:result}
\end{figure*}

@@

This tree shows surprising similarity to the taxonimic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.].
The maize cultivars are all grouped together and there's even the distinction between the non-stiff stalk (W22 and Mo17), iodent (Ph207), and stiff stalk (maize_v4) classes.
Only maize_v3, which is the same cultivar as maize_v4 (B73) is waaaay out there where it probably doesn't belong.
We're currently investigating this but it seems that there is a lot wrong with the genome assembly itself, which would naturally influence the functional annotations derived from it.

@@

```bash
git clone https://github.com/Thyra/EEOB563.git paper_dennis # Clone the repository
cd paper_dennis/final_project # change into the directory
./build_distance_matrix > infile
module load phylip # (if you're on HPC)
neighbor
```

>>>>

`r newthought("When we build phylogenetic trees,")` we often use the sequences of certain genes as the basis for our tree.
They are the direct substrate of evolution so that makes sense, but practically often too abstract to be helpful.
Let's say for example, that you have a patient infected with a novel pathogen.
You can identify that pathogen and place it somewhere in the bacterial taxonomy, but unfortunately, none of the treatments that you usually use for closely related species show any effect against this one.
How come?
Just because two organisms are closely related does not *necessarily* mean that they are similar in their phenotype.
So a tree that groups organisms based on their actual *in vivo* similarity would be much more helpful here.
There are such trees built on phenotypic characteristics but unfortunately the choice of what characteristics to look at is not trivial and always situation-dependent.
They can also be considered quite arbitrary\sidenote{
Look at this (parodic) taxonomy of animals by Jorge Luis Borges:
1. those that belong to the Emperor,
2. embalmed ones,
3. those that are trained,
4. suckling pigs,
5. mermaids,
6. fabulous ones,
7. stray dogs,
8. those included in the present classification,
9. those that tremble as if they were mad,
10. innumerable ones,
11. those drawn with a very fine camelhair brush,
12. others,
13. those that have just broken a flower vase,
14. those that from a long way off look like flies.
}.
What's needed would be something that is exhaustive, clearly defined, and reproducible, but also more meaningful than just a DNA sequence.
And they did it [@Zhu2015]!

In this paper, I'm going to do something slightly similar. Not on bacteria, but on plants though.
I know, disapointing, but I'll find a similarly fascinating reason why that's important soon.
So, to summarize: I want to build a phylogenetic tree on plant species not based on their genetic sequence but on the *functions* that this plant is able to execute.

@@

## Data
Annotating genes with their functions can be done experimentally (e.g. by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try and predict the function of a given gene.
Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner [@Wimalanathan2018].
We have been applying this pipeline to whole-genome assemblies of different plant species and generated functional annotations for every gene in each genome.
These annotation sets are (or will be shortly) available from  
\scriptsize https://dill-picl.org/projects/gomap/gomap-datasets \normalsize.

@@

Applying this formula to all pairwise combinations of the genomes we're looking at yields a $S \times S$ distance matrix that can then serve as the input for a neighbor joining algorithm (provided by PHYLIP).
I rooted the resulting tree manually outside of the grasses (maize, wheat, rice, barley).

@@

The phylogram is displayed in Figure \ref{fig:result}.
4 of the 5 maize assemblies are grouped together in a clade, while the last one is an outlier to the remaining grasses and sits on a very long branch.
In the remaining plants, cotton is an outlier to the legumes, which are all grouped in a single clade.

\begin{figure*}[htbp]
\includegraphics{resources/result.png}
\caption{Resulting phylogram. maize\_v3 and maize\_v4 are two assemblies of the same maize cultivar, medicago1 and medicago2 are two different genotypes.}
\label{fig:result}
\end{figure*}

@@

This tree shows surprising similarity to the taxonimic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.]:
The maize cultivars are all grouped together and there's even the distinction between the non-stiff stalk (W22 and Mo17), iodent (Ph207), and stiff stalk (maize_v4) classes.
Only maize_v3, which is the same cultivar as maize_v4 (B73) is waaaay out there where it probably doesn't belong.
We're currently investigating this but it seems that a lot went wrong with the genome assembly itself, which would naturally influence the functional annotations derived from it.
The two medicago genotypes, while correctly placed together in one clade, are an outlier to the other legumes.
In a taxonomic tree, their place would be switched with peanut.
Steven Cannon, who is an evolutionary biologist working on legumes, assumes that the reason for that is that medicago has a much higher rate of evolution and gene birth and death than other plants [@Young2011].
As you will have noticed, this is all just a work in progress at the moment. My PI and I will meet with Steven on Monday and discuss (among other things) the following questions -- feel free to comment in your review if you have any input!

- How similar is this tree to the taxonomic tree?
- What are the reasons for the differences, what are the reasons for the similarities?
- What biases in the construction process need we be aware of?
- What does the tree depict/how do we interpret it?
- In what way could it be scientifically valuable?
- Are there any other taxa that would be good to have in the tree?
- Is there any further analysis that would be good to do? (e.g. branch support, looking at each GO aspect separately, a more sophisticated measure of GAF similarity, trying a character state based method instead of distance...)
- Is rooting the tree there reasonable?

@@

```bash
git clone https://github.com/Thyra/EEOB563.git paper_dennis # Clone the repository
cd paper_dennis/final_project # change into the directory
./build_distance_matrix > infile
module load phylip # (if you're on HPC)
neighbor # (use standard options)
```
